### 2019-02-25 sections/abstract.tex
abstract: Grounding
examples -> case studies
All changes based on johan's comments

% Grounding
%  ??? De taal zelf is eigenlijk het artifact

% Importance
TOP has been employed in projects with the Dutch coast guard, tax office, and navy.
Our work matters because formal program verification is important for mission-critical software, especially for systems with concurrency.

>>>>

% Grounding
We have implemented the semantic rules of TopHat in Idris, and the task layer on top of iTasks.
This shows that our approach is feasible, and lets us demonstrate the concepts by means of illustrative case studies.

% Importance
TOP has been employed in projects with the Dutch coast guard, tax office, and navy.
Our work matters because Formal program verification is important for mission-critical software, especially for systems with concurrency.

### 2019-02-25 sections/introduction.tex
All changes based on johan's comments

  \item
    Using both the essential concepts and the formal calculus, we compare \TOP with multiple related works, in areas ranging from business process modelling to process algebras and reactive programming.

>>>>

  \item
    Using both the essential concepts and the formal calculus, we compare \TOP with related work, in areas ranging from business process modelling to process algebras and reactive programming.

### 2019-02-25 sections/language.tex
remove old hfill
Fixed minor nit 11 by reviewer #3
small fixes and more fixmes
add stub for abstract example
fixed intro section 4
All changes based on johan's comments

In this section, we present the constructs of the modular interactive workflow
calculus \TOPHAT. We define the host and task language, the types, and
the static semantics. Then we describe the workings of each construct
using examples. These construct are formalised in \cref{sec:semantics}.

@@

Editors stand for various forms of input and output, for example widgets in a \GUI, form fields on a webpage, sensors, or network connections.
%
Consider the editor for a person's age from \cref{exm:flight-booking}.
Users can change the value until they are satisfied with it.
Editors are meant capture this constantly changing nature of user input.
%
The user interface of an editor depends on its type.
This could be an input field for strings, a toggle switch for booleans, or even a map with a pin for locations.
It could also be a parser that tries to parse a line of text to match the type of the editor.

@@

Steps are guarded, which means that the step combinators can only proceed when the following conditions are met.
The left hand side must have a value, only then can the right hand side calculate the successor task.
The successor task must not be $\Fail$, introduced below.
The internal step can proceed immediately when these conditions are met.
The external step must additionally receive a continue event $\Continue$.

@@

\hfill Internal \\ choice ($\Or$) is a parallel \emph{or}.
It picks the leftmost branch that has a value.
Its typing rule \refrule{T-Or} states that both branches must have the same type $\Task \tau$.
For example $\Enter \Int \Or \Edit 37$ normalizes to $\Edit 37$, because $\Enter \Int$ doesn't have a value.
Users can work on both branches of an internal choice simultaneously.

@@

\begin{example}[Delay]

>>>>

In this section, we present the constructs of \TOPHAT, our modular interactive workflow calculus.
We define the host and task language, the types, and the static semantics.
Then we describe the workings of each construct using examples.
These constructs are formalised in \cref{sec:semantics}.

@@

Editors stand for various forms of input and output, for example widgets in a \GUI, form fields on a webpage, sensors, or network connections.
%
Consider the editor for a person's age from \cref{exm:flight-booking}.
Users can change the value until they are satisfied with it.
Editors are meant to capture this constantly changing nature of user input.
%
The user interface of an editor depends on its type.
This could be an input field for strings, a toggle switch for booleans, or even a map with a pin for locations.
It could also be a parser that tries to parse a line of text to match the type of the editor.

@@

Steps are guarded, which means that the step combinators can only proceed when the following conditions are met.
The left hand side must have a value, only then can the right hand side calculate the successor task.
The successor task must not be $\Fail$, introduced below.
This is enforced on the semantic level, as described in the next section.
The internal step can proceed immediately when these conditions are met.
The external step must additionally receive a continue event $\Continue$.

@@

Internal choice ($\Or$) is a parallel \emph{or}.
It picks the leftmost branch that has a value.
Its typing rule \refrule{T-Or} states that both branches must have the same type $\Task \tau$.
For example $\Enter \Int \Or \Edit 37$ normalizes to $\Edit 37$, because $\Enter \Int$ doesn't have a value.
Users can work on both branches of an internal choice simultaneously.

@@

\begin{example}[Delay]
\label{exm:delay}

### 2019-02-25 sections/outlook.tex
change reference to abstract example to breakfast
All changes based on johan's comments

Firstly, we would like to express properties of tasks and prove them.
For example, one would like to prove that, no matter what, in Example~\ref{exm:abstract} an abstract is always submitted.
Secondly, we would like explore what it means for two tasks to be equal.
One could have noticed that some operators have a monadic or applicative feeling.
The combination of $\And$ and $\Edit$ could form a (lax) monoidal functor,
$\Or$ is similar to applicative choice,
and $\Then$ looks like a bind operation.
We need a correct understand of equivalency of tasks,
taking the interactive setting into account,
to prove this.
Thirdly, we do not know yet if the more complex combinators of \ITASKS are expressible in the basic combinators of \TOPHAT.
We implemented \TOPHAT on top of \ITASKS, so we know it is a subset,
but we also know \ITASKS can do more.

>>>>

Firstly, we would like to express properties of tasks and prove them.
For example, one would like to prove that, no matter what, in Example~\ref{exm:breakfast} breakfast is always being served.
Secondly, we would like to explore what it means for two tasks to be equal.
One could have noticed that some operators have a monadic or applicative feeling.
The combination of $\And$ and $\Edit$ could form a (lax) monoidal functor,
$\Or$ is similar to applicative choice,
and $\Then$ looks like a bind operation.
We need a correct understand of equivalence of tasks,
taking the interactive setting into account,
to prove this.
Thirdly, we do not know yet if the more complex combinators of \ITASKS are expressible in the basic combinators of \TOPHAT.
We implemented \TOPHAT on top of \ITASKS, so we know it is a subset,
but we also know \ITASKS can do more.

### 2019-02-25 sections/properties.tex
All changes based on johan's comments

Moving on, we show that normalisation also preserves, by showing that the
following Theorem holds.

@@

Since this semantics makes use of the value function $\Value$ and the striding semantics $\stride$. We first need to
show that preservation also holds for these two semantics.

@@

\begin{lemma}[Striding preserves]
  For all expressions $e$ and states $s$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  if $e,s \stride e',s'$,
  then $\Gamma,\Sigma \infers e':\tau$ and $\Sigma \infers s'$.
  \label{thm:pres-stride}
\end{lemma}

Lemma~\ref{lem:presvalue} and Theorem~\ref{thm:pres-stride} are proven in the
appendix by induction over $e$. This subsequently allows us to prove
Theorem~\ref{thm:pres-norm}, again by induction over $e$.

@@

\begin{theorem}[Preservation under handling]
  % For all well typed expressions $e$, states $s$, and inputs $i$,
  For all expressions $e$, states $s$ and inputs $i$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  if $ e,s \handle{i} e's'$,
  then $\Gamma,\Sigma\infers e':\Task\tau$ and $\Sigma\infers s'$.
  \label{thm:pres-handle}
\end{theorem}

@@

From Theorem~\ref{thm:pres-handle} and Theorem~\ref{thm:pres-norm} we directly
obtain that the driving semantics also preserves.

@@

If an expression $e$ and state $s$ are well-typed, then after normalisation, it
either fails, or there exists some input $i$ that can be handled by it.
In order to prove this Theorem, we require an additional theorem.

@@

In order to validate the function that calculates all possible inputs $\Inputs$,
we want to show that the set of possible inputs it produces is both sound and
complete with respect to the handle semantics. By sound we mean that all inputs
in the set of possible inputs can actually be handled by the handle semantics,
and by complete we mend that the set of possible inputs contains all inputs
that can be handled by the handle semantics. Theorem~\ref{thm:safety-i} expresses
exactly this property.

\begin{theorem}[Inputs function is sound and complete]
  For all well typed expressions $e$, states $s$, and inputs $i$,
  % For all expressions $e$, states $s$, and inputs $i$
  % such that $\Gamma,\Sigma \infers e:\tau$ and $\Sigma \infers s$,
  we have that $i \in \Inputs{(e)}$ if and only if $e,s \handle{i} e',s'$.
  \label{thm:safety-i}
\end{theorem}

>>>>

Moving on, we show that normalisation also preserves types, by showing that the following Theorem holds.

@@

Since this semantics makes use of the value function $\Value$ and the striding semantics $\stride$,
we first need to show that preservation also holds for these two semantics.

@@

\begin{lemma}[Striding preserves]
  For all expressions $e$ and states $s$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  if $e,s \stride e',s'$,
  then $\Gamma,\Sigma \infers e':\Task\tau$ and $\Sigma \infers s'$.
  \label{thm:pres-stride}
\end{lemma}

Lemma~\ref{lem:presvalue} and Lemma~\ref{thm:pres-stride} are proven in the
appendix by induction over $e$. This subsequently allows us to prove
Theorem~\ref{thm:pres-norm}, again by induction over $e$.

@@

\begin{theorem}[Preservation under handling]
  % For all well typed expressions $e$, states $s$, and inputs $i$,
  For all expressions $e$, states $s$ and inputs $i$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  if $ e,s \handle{i} e',s'$,
  then $\Gamma,\Sigma\infers e':\Task\tau$ and $\Sigma\infers s'$.
  \label{thm:pres-handle}
\end{theorem}

@@

From Theorem~\ref{thm:pres-handle} and Theorem~\ref{thm:pres-norm} we directly
obtain that the driving semantics also preserves types.

@@

If an expression $e$ and state $s$ are well-typed, then after normalisation, the pair $e',s'$
either fails, or there exists some input $i$ that can be handled by it under the handling semantics.
In order to prove this Theorem, we require an additional theorem.

@@

In order to validate the function that calculates all possible inputs $\Inputs$,
we want to show that the set of possible inputs it produces is both sound and complete with respect to the handle semantics.
By sound we mean that all inputs in the set of possible inputs can actually be handled by the handle semantics,
and by complete we mean that the set of possible inputs contains all inputs that can be handled by the handle semantics.
Theorem~\ref{thm:safety-i} expresses exactly this property.

\begin{theorem}[Inputs function is sound and complete]
  For all well typed expressions $e$, states $s$, and inputs $i$,
  % For all expressions $e$, states $s$, and inputs $i$
  % such that $\Gamma,\Sigma \infers e:\tau$ and $\Sigma \infers s$,
  we have that $i \in \Inputs{(e,s)}$ if and only if $e,s \handle{i} e',s'$.
  \label{thm:safety-i}
\end{theorem}

### 2019-02-25 sections/relatedwork.tex
All changes based on johan's comments

Much research has been done into workflow modelling. These works focus on
describing the collaboration between subsystems, rather than the communication
between them.
The systems described here follow a \emph{boxes and arrows} model of specifying workflows.
Control flow, represented by arrows, usually can go unrestricted from anywhere to anywhere else in a workflow.
We see \TOP as the functional programming of workflows, as opposed to this \emph{goto}-style.

@@

HipHop \cite{berry2011hiphop,journals/corr/BerryS13} is a programming language tailored to the development synchronous reactive web systems.
From a single source, both server and client applications can be generated.
Programs are written in the Hop language, a Scheme dialect.
Communication is based on a reactive layer embedded in Hop.
The set of HipHop reactive statements is based on those of the Esterel language \cite{boussinot1991esterel,journals/scp/BerryG92}.
Each reactive component starts by specifying possible input and output events.
The component then proceeds as a state machine.

Input events are sent to such a machine programatically using Hop, or are explicitly wired to events from the client.
They are optionally associated with a Hop value.
As Hop is a dynamic language, and HipHop uses strings to identify events, events and their possible associated values are not statically checked.
Events are aggregated until the moment the machine as asked to react.
% then are perceived as conceptually simultaneous.
The machine is executed and reacts by building a multi-set of output events.
The execution of a HipHop machine is atomic.
The set of inputs is not influenced by the current computations.

As with \TOPHAT, HipHop is a \DSL embedded in a general purpose programming language.
% the same way \TOPHAT is embedded into the simply typed $\lambda$-calculus.
Another similarity is that both specifications lead to executable server and client applications from a single source.
However, both HipHop and Esterel are more low level regarding their specification.
Where \TOPHAT takes tasks and collaboration as a starting point,
HipHop focusses on synchronous communication and atom execution of reactive machines.

@@

Another important difference is the way in which both systems handle events.
In HipHop the programmer decides when a machine should processes its events.
This could be just one event, or a multi-set of events that are processed simultaneously.
\TOPHAT always processes an input the moment it occurs and only handles a single event in one instance.

@@

\FRP and \TOP are different systems that have different goals in mind.
Whereas \FRP expresses automatically updating data dependencies, \TOP expresses collaboration patterns.
\TOP has no notion of time.
Tasks can not change spontaneous over time, while behaviours can.
Only input events can change task values.
The biggest conceptual difference between a workflow in \TOP and a data network in \FRP is that an event to a task only causes updates up until the next step, while an event in \FRP propagates through the whole network.

@@

Chorus~\cite{chen2017chorus} is a visual programming environment for online
collaboration apps. The goal of Chorus is to allow users to program their own
application, though which a collective task can be coordinated. Users work on a
preset data-type that drives the behaviour of the user defined applications.

>>>>

Much research has been done into workflow modelling.
This work focusses on describing the collaboration between subsystems,
rather than the communication between them.
The systems described here follow a \emph{boxes and arrows} model of specifying workflows.
Control flow, represented by arrows, usually can go unrestricted from anywhere to anywhere else in a workflow.
We see \TOP as the functional programming of workflows, as opposed to this \emph{goto}-style.

@@

HipHop \cite{berry2011hiphop,journals/corr/BerryS13} is a programming language tailored to the development of synchronous reactive web systems.
From a single source, both server and client applications can be generated.
Programs are written in the Hop language, a Scheme dialect.
Communication is based on a reactive layer embedded in Hop.
The set of HipHop reactive statements is based on those of the Esterel language \cite{boussinot1991esterel,journals/scp/BerryG92}.
Each reactive component starts by specifying possible input and output events.
The component then proceeds as a state machine.

Input events are sent to such a machine programatically using Hop, or are explicitly wired to events from the client.
They are optionally associated with a Hop value.
As Hop is a dynamic language, and HipHop uses strings to identify events, events and their possible associated values are not statically checked.
Events are aggregated until the moment the machine is asked to react.
% then are perceived as conceptually simultaneous.
The machine is executed and reacts by building a multi-set of output events.
The execution of a HipHop machine is atomic.
The set of inputs is not influenced by the current computations.

As with \TOPHAT, HipHop is a \DSL embedded in a general purpose programming language.
% the same way \TOPHAT is embedded into the simply typed $\lambda$-calculus.
Another similarity is that both specifications lead to executable server and client applications from a single source.
However, both HipHop and Esterel are more low level regarding their specification.
Where \TOPHAT takes tasks and collaboration as a starting point,
HipHop focusses on synchronous communication and atomic execution of reactive machines.

@@

Another important difference is the way in which both systems handle events.
In HipHop the programmer decides when a machine should process its events.
This could be just one event, or a multi-set of events that are processed simultaneously.
\TOPHAT always processes an input the moment it occurs and only handles a single event in one instance.

@@

\FRP and \TOP are different systems that have different goals in mind.
Whereas \FRP expresses automatically updating data dependencies, \TOP expresses collaboration patterns.
\TOP has no notion of time.
Tasks cannot change spontaneous over time, while behaviours can.
Only input events can change task values.
The biggest conceptual difference between a workflow in \TOP and a data network in \FRP is that an event to a task only causes updates up until the next step, while an event in \FRP propagates through the whole network.

@@

Chorus~\cite{chen2017chorus} is a visual programming environment for online
collaboration apps. The goal of Chorus is to allow users to program their own
application, though which a collective task can be coordinated. Users work on a
preset data-type that drives the behaviour of the user defined applications.
\fixme{compare with tophat}
